---
title: "Mapping data in R"
author: "Bervelin"
date: 2022-02-25T21:15:14-05:00
categories: ["R", "Articles"]
tags: ["R", "map", "geographic data"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = NA)
```

In this tutorial, we are going to see how to plot data on the map using `sf` and `ggplot2` packages. 

# Introduction

A [geographical] map is a representation of a geographical space. It highlights the extent of this space, its relative location in relation to neighboring spaces, as well as the location of the elements it contains. Maps are also used to represent geographical phenomena, that is to say phenomena whose spatial configuration produces meaning.


```{r message=FALSE, warning=FALSE}
library(sf)      # for the map
library(dplyr)   # for data manipulation (and the pile : %>%)
library(ggplot2) # for plottingNot run)
```


# Loading the shapefile 

We are going to use the shapefile of the Democratic Republic of Congo, which can be found online in the <a href = 'https://geonode.wfp.org/layers/geonode%3Acod_admbnda_adm2_rgc_20170711' target = '_blank'>WFPGeoNode</a> website.
In most cases, the files downloaded will be zipped. After unzipping, store all the files in the same folder.

In our case, our shape files are stored in a folder named _shp_ and contains 6 files. The file we will load ends with _.shp_. 

```{r}
setwd("shp/") 
dir()
```

The following command reads the shape file and stores it in _drc_shape_ object

```{r}
drc_shape <- st_read("shp/cod_admbnda_adm2_rgc_20170711.shp")
```

Let's explore a bit the content of _drc_shape_ with some basic R functions. 

```{r}
class(drc_shape)
```

We can see that the object _drc_shape_ has two classes, _sf_ and _data.frame_. It means that it can be manipulated like standard dataframes.

`str()` function indicates that our data.frame contains 240 observations and 15 varibleas. At the end, we have _geometry_ which contains geographical informations.

```{r}
str(drc_shape)
```

`unique()` with a variable name as an argument gives us names of the 26 provinces of the country (DRC).

```{r}
unique(drc_shape$ADM1_FR)
```

In this article, we are going to work with the Kwango province. So we create a subset of the main dataset and store it in an objet called _kwango_shape_.

```{r}
kwango_shape <- drc_shape %>%
  filter(ADM1_FR == "Kwango")
```

# The first map

Let's create the first map with `ggplot2` and `geom_sf()` from `sf` package. Note that _color_ argument with "blue" as value is used inside `geom_sf()` to change the color of border lines and _fill_ for the color of the map. 

```{r}
ggplot(data = kwango_shape) + 
  geom_sf(color = "blue", fill = "white") + 
  ggtitle(label = "Kwango province map") + 
  theme_void()
```


# Mapping points with longitude and latitude

Suppose that we want to plot gps coordinates on the our previous map to see where the survey took place. For that, we need two columns : _longitude_ and _latitude_. In the following code, we create a sample dataset (survey_df) containing four variables : _longitude_, _latitude_, _status_, _ADM1_FR_ and _num_.

```{r}
# creating a random dataset
set.seed(14) # to make the dataset reproducible
survey_df <- data.frame(longitude = rnorm(20, mean = 18, sd = 0.7),
                        latitude  = rnorm(20, mean = -7, sd = 0.5),
                        status    = sample(c("O", "N"), 20, replace = T),
                        region    = sample(kwango_shape$ADM2_FR, 20, replace = T),
                        num       = rnorm(20, mean = 200, sd = 20)
                     )
```

Let's map now...

```{r}

ggplot(data = kwango_shape) + 
  geom_sf(color = "blue", fill = "white") + 
  geom_point(data= survey_df, aes(x = longitude, y = latitude, color = status), size = 3) + 
  ggtitle(label = "Kwango province map with points") + 
  theme_void()
```


# Thematic map

The following code allows to join two datasets _kwango_shape_ and _survey_sum_ with the `dplyr`'s function `left_join()`. The _by_ argument allows to specify key columns for joining two tables. 

Note that we first calculate the mean value of our numeric variable grouped by region.

```{r}
survey_sum <- survey_df %>%
  group_by(region) %>%
  summarise(num = mean(num))

all_df <- kwango_shape %>%
  left_join(survey_sum, by = c("ADM2_FR" = "region"))
```

To create a thematic map, in the `geom_sf()` function, inside `aes()`, we use the _fill_ argument giving it the name of the variable to illustrate.

```{r}
ggplot() + 
  geom_sf(data =all_df, aes(fill = num), color = "white") + 
  ggtitle(label = "Kwango province Thematic map") + 
#  scale_fill_gradient2("Variable to illustrate", low = "white", high = "blue") + 
  scale_fill_viridis_c("Variable to illustrate") + 
  theme_void()
```

We know that Kwango, Kwilu and Maï-Ndombe were parts of The old Bandundu province before decentralization. Let's build a map containing all the three provinces.

```{r}
ggplot() + 
  geom_sf(data = drc_shape %>% filter(ADM1_FR %in% c("Kwango", "Kwilu", "Maï-Ndombe")), fill = "white") + 
  geom_sf(data =all_df, aes(fill = num), color = "white") + 
  ggtitle(label = "Old Bandundu province") + 
#  scale_fill_gradient2("Variable to illustrate", low = "white", high = "blue") + 
  scale_fill_viridis_c("Variable to illustrate") + 
  theme_void()
```

# Finishing...

Currently R contains tens of packages to work with geographic data. What was shown here is less than 0.25% of what can be done with geographic data in R. For further reading, check the links below 

- _<a href = 'https://geocompr.robinlovelace.net/index.html' target = '_blank'>Geocomputation with R</a>_
- _<a href = 'https://ggplot2-book.org/maps.html' target = '_blank'> ggplot2: Elegant Graphics for Data Analysis</a>_
- _<a href = 'https://ggplot2.tidyverse.org/reference/map_data.html' target = '_blank'>Create a data frame of map data</a>_

All the code of this article and others can be downloaded in <a href = 'https://www.github.com/bervelin-lumesa/r-blog-articles'>Github</a>. 

Was this article helpful ? Don't forget to share it !

_TO KUTANI MBALA YA SIMA :)_


