---
title: "Introduction to Exploratory Data Analysis (EDA) with R"
author: "Bervelin"
date: '2022-02-18T21:13:14-05:00'
categories:
- Articles
- R
tags:
- EDA
- R
- Data analysis
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = NA)
```


In this article, we describe what Exploratory Data Analysis is and see how to perform it with R.


# Introduction to R and Rstudio

<a href = 'https://r-project.org' target = '_blank'>R</a> is an open-source software environment for statistical computing and graphics. R compiles and runs on Windows, Mac OS X, and numerous UNIX platforms (such as Linux).
The R software project was first started by Robert Gentleman and Ross Ihaka. As of 2022, there were more than 3,000 such packages hosted on <a href = 'https://r-project.org' target = '_blank'>CRAN</a>.
R can be installed from the <a href = 'https://r-project.org' target = '_blank'>CRAN</a> website

<a href = 'https://www.rstudio.org' target = '_blank'>RStudio</a> is an integrated development environment (IDE) that adds modern features like syntax highlighting to R. The
strength of RStudio is that it brings all the features that you need together in one place.

# Introduction to EDA

Exploratory data analysis is the process of exploring your data, and it typically includes examining the structure and components of your dataset, the distributions of individual variables, and the relationships between two or more variables. Visualization is the most heavily relied upon tool for exploratory data analysis to get a deep understanding of your data.

Exploratory analysis helps us ask better questions, but it does not answer questions. More specifically, we explore data in order to:

- Understand data properties such as nonlinear relationships, the existence of missing
values, the existence of outliers, etc.
- Find patterns in data such as associations, group differences, confounders, etc.
- Suggest modeling strategies such as linear vs. nonlinear models, transformation

In the <a href = 'https://leanpub.com/tidyverseskillsdatascience' target = '_blank'>Tidyverse Skills for Data Science in R</a>, authors summarized the general principales of EDA as follows : 

- Look for missing values
- Look for outlier values
- Use plots to explore relationships
- Use tables to explore relationships
- If necessary, transform variables

> A detective investigating a crime needs both tools and understanding. If he has no fingerprint powder, he will fail to find fingerprints on most surfaces. If he does not understand where the criminal is likely to have put his fingers, he will not look in the right places. Equally, the analyst of data needs both tools and understanding. - _John W. Tukey_

As stated in the quote above, doing Exploratory Data analysis requires to have tools and understanding. In our case, the R language with all its useful functions can be considered as our tool and the domain knowledge as our understanding. 

# Practice

## Data

Data used in this post are from the `questionr` package. hdv2003 is a sample from 2000 people and 20 variables taken from the _Histoire de Vie survey_, produced in France in 2003 by INSEE. `questionr` package can be installed as other packages by typing : `install.packages("questionr")` and load to your RStudio session with the following command : `library(questionr)`. Finally, `data(hdv2003)` allows to access data.

Apart from the `questionr` package where our data come from, we will also use  some functions from the `tidyverse` package for data minipulation and graphics and the `naniar` package for missing data visualization. So you need to do the same work as with `questionr` package, except the last command.

## EDA with R

First of all, we need to load packages we'll work with and our data.
Note that our `data.frame` is transformed to a `tibble` with `as_tibble()` for convenient display of our data.

```{r message=FALSE, warning=FALSE}
# libraries
library(tidyverse)
library(questionr)
library(naniar)

# data
data(hdv2003) 

hdv2003 <- as_tibble(hdv2003)
```

The following command allows us to know the dimension of our dataset.

```{r}
dim(hdv2003) 
```

Our dataset contains 2000 observations and  20 variables.

`names()` or `colnames()` returns the variable names of the dataset.

```{r}
names(hdv2003) 
```

Now, we would like to see the structure of our dataset. The `str()` gives the data type of variables and a look at some rows.

```{r}
str(hdv2003) 
```

We can see that some variables are numeric, integer, while others are factors (lables with numeric values associated). If you think that a certain variable has the wrong type, functions like `as.numeric()`, `as.factor()`, `as.Date()` can be used to fix.

The following output uses `head()` and `tail()` to print the first and the last six rows of our dataset.

```{r}
head(hdv2003) 

tail(hdv2003) 
```

To get a statistical summary of our data, we use the `summary()` function which gives the minimum, first and third quantiles, mean, median and the maximum values of all numeric variables. For categorical variables (factor), it counts the number of observations within each category. 

```{r}
summary(hdv2003)
```

You may also use `glimse()` function from the `skimr()` package like this `skimr::skim(hdv2003)`

We see for instance that the mean age is 48.16, with minimum of 18 and 97 as maximum. Our data contains more women (1101) than men (899).

`summary()` can also be used to describe a single variable, while `table()` or `freq()`  (from `questionr`) is used for a categorical variable.

```{r}
summary(hdv2003$age)
```


```{r}
freq(hdv2003$relig)
```


From the output of the first summary function call, we notice that some variables have many missing data. For instance, _qualif_ has 347 missings  whereas _trav.satisf_ and _trav.imp_ have 952 missing values.
We can use the `vis_miss()` function from the `naniar` package the visualize those missing values.

```{r}
vis_miss(hdv2003)
```

Our dataset contains 9.9% of missings values, and those missings mostly come from _trav.imp (47.6%)_, _trav.satisf (47.6%)_, _qulif (17.4%)_ and _nivetud (5.6%)_.

`naniar` has other useful functions to handle missing data, such as `n_var_miss()` which gives you the number of variables containing missing values.

Sometimes we want to see the statistics broken down by a categorical variable. For that, we can use the powerful functions from `dplyr` packages, which comes with `tidyverse`.

```{r}
hdv2003 %>%
  group_by(trav.satisf) %>%
  summarise(min    = min(heures.tv, na.rm = T),
            mean   = mean(heures.tv, na.rm = T),
            median = median(heures.tv, na.rm = T),
            max    = max(heures.tv, na.rm = T))
```


Histograms can help to identify the shape of the distribution of a variable. 

```{r histogram, message=FALSE, warning=FALSE}
ggplot(data = hdv2003) + 
  geom_histogram(aes(x = hdv2003$age))
```

We can see that values are around 50 years and the presence of small queue in the right side.

`ggplot2` allows to facet a plot with a categorical variable. To facet with _religion_, we add `facet_wrap(~ relig)` to the previous plot.

```{r histogram_facet, message=FALSE, warning=FALSE}
ggplot(data = hdv2003) + 
  geom_histogram(aes(x = hdv2003$age)) + 
  facet_wrap(~ relig)
```

Boxplots can also help to study the distribution, but also for identify outliers (very small or big values). 

```{r boxplot, message=FALSE, warning=FALSE}
ggplot(data = hdv2003) + 
  geom_boxplot(aes(x = occup, y = heures.tv, color = occup)) + 
  theme(legend.position = "none")
```

We can also check for relationships among numeric variables using scatterplots.

```{r}
ggplot(data = hdv2003) + 
  geom_point(aes(x = age, y = poids, color = sexe)) 
```

Exploratory Data Analysis can also suggest to recode some variables and or make transformations...

Here we create a new variable _minutes.tv_ which gives the time of watching tv in minutes, using `mutate()`.

```{r message=FALSE, warning=FALSE}
hdv2003 %>%
  mutate(minutes.tv = heures.tv * 60) %>% 
  select(heures.tv, minutes.tv)
```

# Finishing...

We have seen how to perform EDA using some functions of the R language. Summaries, tables and graphics were used to describe data. Results from an Exploratory Data Analysis can help to further with Inferential analysis. For instance, we can test if there is a statistical difference in the time of watching TV among different occupation (occup) categories.

For more reading, check the following links : 

- _<a href = 'https://r4ds.had.co.nz/exploratory-data-analysis.html' target = '_blank'>R for Data Science</a>_
- _<a href = 'https://www.pluralsight.com/guides/exploratory-data-analysis-in-r' target = '_blank'>Exploratory Data Analysis in R with Tidyverse</a>_
- _<a href = 'https://bookdown.org/rdpeng/exdata/' target = '_blank'>Exploratory Data Analysis with R</a>_

All the code of this article and others can be downloaded in <a href = 'https://www.github.com/bervelin-lumesa/r-blog-articles'>Github</a>. 

Was this article helpful ? Don't forget to share it !

_TO KUTANI MBALA YA SIMA :)_

